---
title: "GroupProjectQuarto"
author: "Claire Savage, Dani Wells,Abby drongpa"
format: html
editor: visual
self-contained: true
---

```{r setup}
library(tidyverse)
library(here)
```

## Data and Variable Descriptions

For this project we will be looking at the Relationship between life expectancy at birth and total CO2 emissions. We will be using the "Life expectancy, at birth" and "Consumption CO2 emissions, million tonnes" variables from Gapminder in our analysis. Life expectancy at birth is measured in years and measures the number of years an infant born in the given country in the given year would live if mortality rates stay the same throughout its life. Total CO2 emissions is measured in million tonnes and is the total carbon dioxide emitted from each country per Capita.

## Hypothesized Relationship

Here we describe what we think the relationship is and put any outside sources we used

## Data Cleaning

#Claire

```{r}
1+1
2*2
```

#Abby

```{r}

```

Pivot Longer

Clean Data

Joining Data
