---
title: "GroupProjectQuarto"
author: "Claire Savage, Dani Wells, Abby drongpa"
format: html
editor: visual
code-fold: true
self-contained: true
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(here)
CO2_emissions <- read_csv(here::here("co2_cons.csv"))
```

## Data and Variable Descriptions

For this project we will be looking at the Relationship between life expectancy at birth and total CO2 emissions. We will be using the "Life expectancy, at birth" and "Consumption CO2 emissions, million tonnes" variables from Gapminder in our analysis. Life expectancy at birth is measured in years and measures the number of years an infant born in the given country in the given year would live if mortality rates stay the same throughout its life. Total CO2 emissions is measured in million tonnes and is the total carbon dioxide emitted from each country per Capita.

## Hypothesized Relationship

Here we describe what we think the relationship is and put any outside sources we used

## Data Cleaning

#Claire

```{r}
CO2_emissions_clean <- CO2_emissions %>% 
  mutate(across(`1800`: `2022`, as.numeric)) #%>% 
  #pivot_longer(cols = `1800`:`2022`,
   #            names_to = "year",
    #           values_to = "emissions")
# the negative symbols in the data set, can use str(CO2_emissions) to see which years have non-numeric values/something that is making them be considered a character (ie this one has some weird negatives)
```


#Abby

```{r}
Lex_data <- read.csv(here::here("lex.csv"))
```

Pivot Longer

Clean Data

Joining Data
