---
title: "GroupProjectQuarto"
author: "Claire Savage, Dani Wells, Abby Drongpa"
format: html
editor: visual
code-fold: true
self-contained: true
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(here)
library(broom)
library(patchwork)
library(DT)
library(gganimate)
library(gifski)
CO2_emissions <- read_csv(here::here("co2_cons.csv"))
lex_data <- read_csv(here::here("lex.csv"))
```

## Introduction

As the world experiences the increasing effects of climate change, it becomes important to look at how different aspects of climate change affect our lives. For this project we looked at the relationship between life expectancy at birth and total CO2 emissions. In order to do this, we used the “Life expectancy, at birth” and “Consumption CO2 emissions, million tonnes” variables from Gapminder in our analysis, with both of these data sets including information on 186 different countries.

Total CO2 emissions was the explanatory variable, and was measured in million tonnes. It is total carbon dioxide emitted from each country per Capita. Life expectancy at birth was the response variable, and it was measured in years. It is the number of years an infant born in a given country in a given year would live if mortality rates stay the same throughout its life. We chose to study the relationship between these variables over the years 1920 to 2019 because these years were the most recent for which most countries had data available. We chose to omit the data post 2019 because we thought the COVID-19 pandemic might have altered the trends visualized in a way which is not representative of the data set as a whole.

We believe that increased levels of CO2 emission decrease life expectancy, which is line with a study conducted by the [National Library of Medicine](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8782287/#:~:text=Carbon%20emissions%20have%20a%20significantly,decreases%20life%20expectancy%20by%200.012%25.) which found that, “Carbon emissions have a significantly negative impact on life expectancy, suggesting that higher the carbon emissions lower the life expectancy.”

## Methods

In order to analyze the data sets, we started by cleaning them with tidyverse. For the CO2 emissions data set, the negative values in Panama were encoded dashes, so we replaced them with regular dashes in order to convert those numbers to numeric values. The years in question, 1920-2019 were then selected and the table was pivoted to have each year and emission value combination with its own row. For the life expectancy data, we first selected the same years as the CO2 emissions (1920-2019), and then took out any NA values. We then pivoted the table so that it was in a similar format to the CO2 emissions, with each year and life expectancy combination in its own row. Once the individual data sets were properly formatted, we combined them so that for each country, there was a different row with the respective CO2 emissions and life expectancy values.

```{r}
CO2_emissions_clean <- CO2_emissions %>%
  mutate(across(c(`2003`:`2006`,`2011`:`2013`), 
                ~str_replace(.x, 
                             pattern = "−", 
                             replace = "-")), 
         across(c(`2003`:`2006`,`2011`:`2013`), ~
                  as.numeric(.x))) %>% 
  select(country, `1920`:`2019`) %>% 
  pivot_longer(cols = `1920`:`2019`,
               names_to = "year",
               values_to = "emissions")

lex_clean <- lex_data %>%
  select(country, `1920`:`2019`) %>%
  filter(if_all(c(`1920`:`2019`), ~ !is.na(.))) %>%
  pivot_longer(cols = `1920`:`2019`,
               names_to = "year",
               values_to = "life_exp")

co2life_joined <- inner_join(CO2_emissions_clean, lex_clean)

#Seperating the data into 20 year chunks
years_1920_1939 <- as.character(seq(1920, 1939))
years_1940_1959 <- as.character(seq(1940, 1959))
years_1960_1979 <- as.character(seq(1960, 1979))
years_1980_1999 <- as.character(seq(1980, 1999))
years_2000_2019 <- as.character(seq(2000, 2019))

co2_life_twodecade <- co2life_joined|>
  mutate(
    year = as.factor(year)
  )|>
  mutate(
    twodecade = fct_collapse(year, 
                             `1920-1940` = years_1920_1939,
                             `1940-1959` = years_1940_1959,
                             `1960-1979` = years_1960_1979,
                             `1980-1999` = years_1980_1999,
                             `2000-2019` = years_2000_2019
    )

  ) 


#Averaging the data by country 
avg_co2_life_twodecade <- co2_life_twodecade |>
  group_by(country, twodecade) |>
  summarise(
    avg_emissions = mean(emissions, na.rm = TRUE),
    avg_life_exp = mean(life_exp, na.rm = TRUE)
  )

```

## Results

To visualize the relationship between CO2 emissions and life expectancy at birth we used a scatter plot to map the relationship between the variables overtime. For readability we separated the data into five twenty year periods in which we related life expectancy on the Y-axis to CO2 emissions on the X-Axis. The plot was animated to cycle between the 20 year periods using gganimate.

REMEMBER TO PUT THE GG PLOT BACK IN

From this graph we observed that CO2 emissions and life expectancy at birth have increased for most most countries. We also see that life expectancy tends to start plateauing around the 80-85 year range, while CO2 emissions continue to grow exponentially for some countries. We also saw that there is not a strong relationship between life expectancy and CO2 emissions as there is not one strict line that we observed in the scatter plot. However if we were to classify a possible relationship between life expectancy and C02 emissions based on these graphs then it would be a weak positive correlation with numerous outliers that seem to follow the trend of exponential growth of CO2.

#### Linear Regression

To get a clearer picture of how life expectancy relates to CO2 emissions across countries, we took the average of both Life Expectancy and CO2 emissions for all years for each country and created a scatter plot of values. We also fitted a linear regression line to show the relationship we attain when regressing the average life expectancy of each country on it's average CO2 emissions.

```{r}
#The relationship between the two quantitative variables you are investigating.
avg_co2_life <- co2life_joined |>
  group_by(country) |>
  summarise(
    avg_emissions = mean(emissions, na.rm = TRUE),
    avg_life_exp = mean(life_exp, na.rm = TRUE)
  )
```

```{r}
avg_co2_life %>%
  ggplot(aes(x = avg_emissions, y = avg_life_exp)) + 
  geom_point(size = 0.8, alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() + 
  labs(
    title = "Scatterplot of CO2 Emissions vs. Life Expectancy over Time",
    subtitle = "Average Life Expectancy (years)",
    x = "Average CO2 Emissions (million tonnes per capita)",
    y = "") +
  theme(
    plot.title = element_text(hjust = 0.5)
    )

influential_outliers <- avg_co2_life|>
  filter(avg_emissions > 1000)

datatable(influential_outliers, class = 'cell-border stripe hover order-column',
          options = list(searching = FALSE, 
                         lengthMenu = list(c(10, -1), c('10', 'All')),
                         paging = FALSE), 
          colnames = c("Country" = "country", 
                       "Average Emissions" = "avg_emissions",
                       "Average Life Expectancy" = "avg_life_exp"),
          caption = "Table 1: Influential Outliers"
)

```

Based on this scatterplot we again observed that the relationship between life expectancy and CO2 emissions is very weak. However our linear regression showed a positive correlation between life expectancy and CO2 emissions, which is likely due to the extreme outliers in our data. For example, China and the United States have extremely high levels of average CO2 emissions and average life expectancy which would bias our estimate of the relationship between average life expectancy and average CO2 emissions.

```{r}
mod_co2life <- lm(avg_life_exp ~ avg_emissions, data = avg_co2_life)
tidy_co2life <- tidy(mod_co2life)

tidy_co2life <- tidy_co2life|>
  mutate(
    across(estimate:p.value, ~round(.x, 3))
  )

datatable(tidy_co2life,  class = 'cell-border stripe hover',
          options = list(searching = FALSE, 
                         lengthMenu = list(c(10, -1), c('10', 'All')),
                         paging = FALSE),
          colnames = c("Coefficient Estimate" = "estimate", 
                       "Standard Error" = "std.error",
                       "T-Statistic" = "statistic",
                       "P-Value" = "p.value"),
          caption = "Table 2: Estimated Coefficients of the Linear Regression",
)
```

From our linear regression of average life expectancy on average CO2 emissions we get the equation: $\widehat {Avg\,Life\,Expectancy} = 56.41 + 0.01(Avg\,Emissions)$

Based on this equation, the predicted average life expectancy of someone living in a country with no CO2 emissions would be approximately 56.41 years. Additionally, the coefficient on average emissions of 0.01 means that for each additional 100 million tonne increase in average CO2 emissions per capita, predicted average life expectancy increases by 1 year.

#### Model Fit

```{r}
modfit_table <- augment(mod_co2life) %>% 
  summarize(life_re = var(avg_life_exp), 
            life_fitt = var(.fitted),
            life_resid = var(.resid)
            )

modfit_table <- modfit_table|>
  mutate(
    across(life_re:life_resid, ~round(.x, 3))
  )

datatable(modfit_table,  class = 'cell-border stripe hover',
          options = list(searching = FALSE, 
                         lengthMenu = list(c(10, -1), c('10', 'All')),
                         paging = FALSE),
          colnames = c("Variance in Life Expectancy" = "life_re", 
                       "Variance in Fitted Values" = "life_fitt",
                       "Variance in Residuals" = "life_resid"),
          caption = "Table 3: Model Fit Variances",
)
```

There is a lot of variability in the actual average life expectancy values, which had a value of about 206.51. This means that the average observed life expectancy in the data set was extremely spread out. Since the variance in the model predictions is much lower than the variance in the response variable, it means that the model is not accurately explaining a large proportion of the variability. In other words, CO2 emissions explain very little of the variability in average life expectancy. Lastly, the high variance of the residuals means that the residuals are largely spread out, and the relationship is likely not linear.

Based on the model, there is not a linear relationship between average CO2 emissions and average life expectancy. Additionally based on the high variability of the residuals, the conditions of normality and equal variance of the residuals is likely not met. However since the countries were all sampled separately, there is independence between observations. Another way to check if it is a good model is through predictive checks.

#### Predictive checks

In order to measure the fit of our model to the data set we observe from Gapminder we can generate simulated data to compare to our observed. If there were a simple linear relation between average life expectancy and CO2 emissions we would assume that we could reproduce our observed data by predicting values from a normal fitted function and adding in normally distributed errors. Thus, we simulated a random normal distribution of data with the same standard deviation that we observed from our data set and then compared our observed data to our simulated data using two side by side scatter plots.

```{r}
mod_co2life_predict <- predict(mod_co2life) 

mod_co2life_sigma <- sigma(mod_co2life)

noise <- function(x, mean = 0, sd) {
  x + rnorm(length(x),
            mean, 
            sd)
}

sim_response <- tibble(sim_life_exp = noise(mod_co2life_predict,
                                            sd = mod_co2life_sigma))

sim_data <- avg_co2_life %>% 
  filter(!is.na(avg_life_exp),
         !is.na(avg_emissions)) %>% 
  cbind(sim_response)
```

```{r, fig.align = "center", fig.width = 8, fig.cap = "Comparison of Observed vs Simulated values for Life Expectancy Depending on CO2 Emissions."}
obs_reg_avgco2life <- avg_co2_life %>%
  ggplot(aes(x = avg_emissions, y = avg_life_exp)) + 
  geom_point(size = 0.8, alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() + 
  labs(
    title = "Scatterplot of Observed Avg. Life \nExpectancy vs. Avg. CO2 Emissions",
    subtitle = "Average Life Expectancy (years)",
    x = "Average CO2 Emissions (million tonnes per capita)",
    y = "") +
  theme(
    plot.title = element_text(hjust = 0.5)
    )

sim_reg_avgco2life <- sim_data %>%
  ggplot(aes(x = avg_emissions, y = sim_life_exp)) + 
  geom_point(size = 0.8, alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() + 
  labs(
    title = "Scatterplot of Simulated Avg. Life \nExpectancy vs. Avg. CO2 Emissions",
    subtitle = "Simulated Average Life Expectancy (years)",
    x = "Average CO2 Emissions (million tonnes per capita)",
    y = "") +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

obs_reg_avgco2life + sim_reg_avgco2life
```

We created these side-by-side comparison using the [patchwork package](https://github.com/thomasp85/patchwork). From this comparison we notice a few differences between our observed and simulated data, the biggest one being the concentration of data points around the y axis. In our observed data that have around 0 carbon emissions are fairly evenly distributed between 40 and 70 years of age, where as in our simulated data we see more of what we expect from a normal distribution where the simulated data points are fairly centered around an average life expectancy of 55 years and seems normally distributed on either side. 

#### Multiple Predictive Checks

Once we had the side-by-side plots of the observed and simulated data, ran multiple predictive checks to see if the model was a good fit for our data.

```{r}
simulate_and_regress <- function(mod_co2life_predict, mod_co2life_sigma, avg_co2_life) {
  # Generate sim data
  sim_response_rep <- tibble(sim_life_exp = noise(mod_co2life_predict, sd = mod_co2life_sigma))
  
  # Combine with observed data
  sim_data_rep <- avg_co2_life %>% 
    filter(!is.na(avg_life_exp), 
           !is.na(avg_emissions)) %>% 
    cbind(sim_response_rep)
  
  # Perform linear regression and save R-squared
  lm_model <- lm(avg_life_exp ~ sim_life_exp, data = sim_data_rep)
  rsquared <- summary(lm_model)$r.squared
  
  return(rsquared)
}

```

```{r}
comparison_mod <- lm(avg_life_exp ~ sim_life_exp, data = sim_data)
comparison_rsquared <- summary(comparison_mod)$r.squared

# Set seed for reproducibility
set.seed(2003)

r_squared_values <- map_dbl(1:1000, ~simulate_and_regress(mod_co2life_predict, mod_co2life_sigma, avg_co2_life))

r_squared_values %>%
  tibble(rsquared = r_squared_values)%>%
  ggplot(
    aes(x = rsquared)
  ) +
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black") +
  labs(title = "Distribution of R-Squared Values for Predictive Checks",
       x = "R-Squared",
       y = "",
       subtitle = "Frequency") + 
  theme_bw()
```

## Discussion

```{r}
sim_dataround <- sim_data|>
  mutate(
    across(avg_emissions:sim_life_exp, ~round(.x, 3))
  )
  
datatable(sim_dataround,  class = 'cell-border stripe hover order-column',
          colnames = c("Country" = "country", 
                       "Average Emissions" = "avg_emissions",
                       "Observed Average Life Expectancy" = "avg_life_exp",
                       "Simulated Average Life Expectancy" = "sim_life_exp"),
          caption = "Table 2: A comparison between the Observed data from Gapminder and the data we have simulated",
          filter = "top"
          )

```
From our data set, we found that a linear regression was not a good model to use. Not only did the original data not meet the validity conditions for it, but when the simulated data was created based on the model it showed that the regression model did not accurately predict average life expectancy. Additionally, after running multiple predictive checks and recording the $R^2$, it was seen that a majority of the simulations had an $R^2$ of 0, meaning that 0% of the variability in average life expectancy was explained by the average CO2 emissions per capita.  

Since the regression model did not fit the data set, it would be interesting to see what other models might better predict or account for the variance in average life expectancy. It should also be acknowledge though, that there are many different factors that influence the life expectancy of people, and it might be a very complicated model to try to build.


## Conclusion
